The Cost of Exclusion: Political Discrimination and Trust Deficits in the European Union

### 2025-07-07 article/main.tex
Literature review and Discussion added to article

@@

@@

However, due to missing values across the variables of interest, the effective analytical sample for this study is smaller than the total sample published by The World Justice Project. Table \ref{table:1} presents descriptive statistics for the key demographic and political variables among survey respondents included in the study \emph{full} sample.

@@

Researchers typically turn to randomized controlled experiments—the gold standard for causal inference—to address such questions. However, the nature of our research question makes it both practically difficult and ethically problematic to randomly expose individuals to discrimination or harassment based on their political views. As a result, we adopt quasi-experimental designs to examine the causal effects of political discrimination and harassment on institutional trust [citations needed?].

@@

Finally, we investigate the mechanisms through which political discrimination and harassment affect trust in political institutions. We explore two potential pathways: political polarization and political alignment.

Regarding political polarization, we hypothesize that individuals who experience D/H events may become more politically polarized, which in turn could lead to lower levels of trust in political institutions. To test this hypothesis, we include a measure of political polarization in our regression model and assess the transmission mechanism using the mediation analysis proposed by Imai et al. (2011) and Imai and Yamamoto (2013).\footnote{We use the mediation R package developed by Tingley et al. (2014) for the practical implementation of this framework.}

The mediation analysis follows a two-step process. First, we examine how D/H experiences affect political polarization (see Equation \ref{eq:2}). Second, we analyze how both D/H experiences and political polarization jointly influence trust in political institutions (see Equation \ref{eq:3}). We measured political polarization at the individual level using a measure of political conformity/dissonance based on the political ideologies of the respondents. This measure is calculated as the square of the difference between the respondent's political ideology and the average political ideology of the country.\footnote{It is important to note that we use the full study sample to estimate the average political ideology of the country, rather than the matched sample. We believe that using the full representative country sample captures the true distance between the respondent's political ideology and the average political ideology of the country.}

\begin{equation}
\label{eq:2}
M_{ic} = \alpha + \phi X_{ic} + \mathbf{Z}_{ic}'\mathbf{\beta} + \gamma_c + \epsilon_{ic}
\end{equation}
\begin{equation}
\label{eq:3}
Y_{ic} = \alpha + \delta X_{ic} + \rho M_{ic} + \mathbf{Z}_{ic}'\mathbf{\eta} + \lambda_c + \mu_{ic}
\end{equation}

Where:
\begin{itemize}
  \item \( Y_{ic} \): political trust for individual \( i \) in region \( c \)
  \item \( X_{ic} \): D/H event (1 = treated, 0 = control)
  \item \( M_{ic} \): political dissonance for individual \( i \) in region \( c \)
  \item $\phi$, $\delta$, and $\rho$: coefficients of interests
  \item \( \mathbf{Z}_{ic} \): vector of covariates
  \item $\mathbf{\beta}$ and $\mathbf{\eta}$: coefficient vectors for covariates
  \item $\gamma_c$ and $\lambda_c$: region fixed effects
  \item $\epsilon_{ic}$ and $\mu_{ic}$: error terms
\end{itemize}

After fitting these models, we use bootstrapping to estimate causal effects under counterfactual scenarios. In each simulation, we compare predicted trust levels between two scenarios: experiencing D/H events versus not experiencing them, while allowing political dissonance to adjust to its expected level under each condition. Averaging across simulations yields the average causal mediation effect (ACME), which captures the indirect effect through polarization, and the average direct effect (ADE), which captures the remaining direct impact.

@@

\begin{equation}
  \label{eq:4}
  log(Y_{ic}) = \alpha + \tau X_{ic} + \ \mathbf{Z}_{ic}'\mathbf{\beta} + \gamma_c + \varepsilon_{ic}
  \end{equation}

@@

\subsection{Polarization as a transmission channel}

As mentioned earlier, we hypothesize that political polarization serves as a transmission channel through which political discrimination and harassment affect trust in political institutions. To test this hypothesis, we conduct a mediation analysis using the framework developed by Imai et al. (2011) and Imai and Yamamoto (2013). This analysis aims to determine how much of the total effect of D/H experiences on institutional trust operates through political polarization. The framework is based in the estimation of the system presented by equations \ref{eq:2} and \ref{eq:3} along with a bootstrapping procedure to test the mediation channel.

Table \ref{table:6} presents three key estimates: the total effect, the average causal mediation effect (ACME), and the average direct effect (ADE). As we can observe, experiencing political discrimination or harassment leads to a large and highly significant drop in institutional trust: the Total Effect of –0.863 (on a 0–10 scale) tells us that, on average, those who report discrimination score 0.863 points lower in trust than those who do not. This effect is equivalent to the 18.9\% decrease presented in Table \ref{table:4}, column (III).

To better understand the mechanisms behind this decline in trust, the analysis separates the effect into two components: the part that operates indirectly through increased political dissonance (the Average Causal Mediation Effect, or ACME), and the part that operates directly, independently of dissonance (the Average Direct Effect, or ADE).

The estimated ACME is –0.0063, meaning that only about 0.006 points of the decline in trust can be attributed to a pathway where discrimination increases political dissonance, which in turn lowers trust. By contrast, the ADE is –0.857, nearly identical to the total effect, which shows that the vast majority of the impact of discrimination on trust occurs directly, without being mediated by political dissonance.

In other words, less than 1\% of the total effect of political discrimination on institutional trust can be explained by changes in political dissonance. Although this indirect effect is statistically significant, it is substantively negligible. This implies that political dissonance does act as a mechanism, but it plays only a very minor role. Most of the loss in trust stems from other pathways—perhaps emotional responses like anger or fear, cognitive assessments of systemic injustice, or broader alienation from the political system that are not captured by the dissonance variable.

\input{"tabs/mediation_analysis.tex"}

\subsection{Heterogenous effects by political alignment}

@@

>>>>

While Europe has historically exhibited lower levels of affective polarization and declining partisan identification, recent studies suggest the presence of partisan hostility and negative feelings driven by factors such as ideological distance, perceived party threats, and regional tensions. Yet, evidence on longitudinal trends is still inconclusive. Scholars continue to emphasize the need for improved measurements and long-term data to determine whether affective polarization is indeed increasing across European democracies (Wagner, 2024).

Importantly, the rise of far-right populist parties has introduced a distinct affective dimension to European politics, one marked by widespread hostility toward political out-groups. This emerging divide signals the presence of a particularly polarized and emotionally charged layer of political competition (Reiljan, 2019; 2025).

Amid this environment, institutional trust is often viewed as a background condition for democratic stability. However, trust plays a more active and central role—it serves as a fundamental mechanism through which political systems gain legitimacy and foster public compliance. A robust body of literature supports the idea that institutional trust enhances democratic resilience, encouraging civic engagement and reinforcing institutional performance (Marien, 2011; Devine and Valgarosson, 2024).

This paper introduces the concept of grievance-based trust erosion, proposing that trust in institutions can deteriorate when individuals accumulate experiences of perceived injustice, exclusion, or marginalization. While prior research has extensively examined economic, ethnic, and cultural grievances, far less attention has been paid to the role of political exclusion—particularly how discrimination or harassment based on political beliefs may erode trust in public institutions.

To address this gap, we investigate whether experiencing political discrimination or harassment—what we refer to as D/H experiences—has a causal effect on trust in political institutions. This question is not trivial. Individuals who report experiences of political discrimination may differ systematically from those who do not in ways that also affect their level of institutional trust. If such differences are not properly accounted for, naive comparisons may lead to biased or misleading conclusions about the relationship between discrimination and trust.

To overcome this challenge, we employ matching methods and sensitivity analyses to estimate the causal effect of D/H experiences on institutional trust. Our analysis draws on a novel dataset from the EUROVOICES project, a general population survey carried out by the World Justice Project across all 27 European Union Member States. The dataset offers statistically representative coverage across 110 subnational regions, providing an unprecedented level of geographic and demographic granularity.

Our findings reveal strong and robust evidence that political discrimination and harassment significantly erode trust in institutions. Individuals who report such experiences exhibit trust levels that are 16\% to 19\% lower than those who have not experienced discrimination or harassment.

However, the effect of discrimination is not uniform—it varies depending on the individual's political alignment and sense of political dissonance. We show that individuals who experience discrimination and simultaneously feel ideologically out of place experience the largest declines in institutional trust. Although trust decreases across the board, the magnitude of the effect differs significantly between politically aligned and non-aligned individuals. Among those not aligned with the incumbent party, experiencing D/H is associated with a 20\% reduction in trust, compared to a 14\% reduction for aligned individuals.

Furthermore, among individuals with low political dissonance, D/H experiences do not significantly affect trust if the individual is politically aligned. In contrast, non-aligned individuals in this group experience an 18\% drop in trust. On the other hand, for individuals experiencing high political dissonance, trust erosion remains high—between 18\% and 21\%—regardless of political alignment. Notably, the difference between aligned and non-aligned individuals in this group is not statistically significant.

Taken together, these findings underscore a layered erosion of institutional trust in politically polarized settings. Rather than unfolding through a singular mechanism, the effects of political discrimination operate through interlinked and mutually reinforcing pathways—ideological alienation and partisan exclusion—that deepen perceptions of institutional unfairness.

This study makes three key contributions to the literature on institutional trust. First, it broadens existing research by placing political discrimination—an often neglected dimension compared to ethnicity or migration status—at the center of the analysis. Second, it explores the politicization of institutional trust as a mechanism through which political exclusion leads to declining confidence in public institutions. In polarized contexts, individuals increasingly interpret institutions through partisan lenses, with those experiencing political discrimination more likely to view the state as aligned with opposing political factions. Third, by empirically testing this causal chain at the subnational level across Europe, the study provides novel evidence on how localized experiences of political exclusion and emotional polarization interact to shape public attitudes toward democratic institutions.

The remainder of this article is organized as follows. Section 2 provides a brief literature review outlining the current state of research on the relationship between trust in public institutions and political discrimination, with particular attention to the factors shaping this relationship in the context of growing political polarization. This section also introduces our central research hypothesis. Section 3 describes the dataset used in this study—a novel subnational survey covering 110 regions across the European Union. Section 4 presents an exploratory analysis of how political discrimination manifests in the data and what patterns it reveals. In Section 5, we detail the methodological approach employed to test our hypothesis, including the matching strategy and sensitivity analysis. Section 6 presents our main empirical findings. Finally, Section 7 discusses the broader implications of our results and offers concluding remarks.

@@

Institutional trust is often treated as a background condition for democratic stability. Yet, rather than serving as a passive backdrop, it functions as a central mechanism through which political systems earn legitimacy and secure compliance. Democratic institutions rely not only on performance or coercion but on public belief in the fairness of rules, the impartiality of procedures, and the integrity of leaders (Levi \& Stoker, 2000; Citrin \& Stoker, 2018).

Trust in institutions reflects more than evaluations of policy outcomes—it signals deeper commitments to institutional principles and the constitutional order (Easton, 1965). Building on this foundation, scholars have shown that institutional trust reinforces civic engagement and perceptions of procedural fairness (Tyler, 2006). More recent work highlights its crucial role in promoting democratic resilience, particularly during crises or episodes of institutional contestation (Zmerli \& Van der Meer, 2017; Devine \& Valgarosson, 2024).

Empirical evidence supports the view that trust strengthens democracy. For instance, Marien (2011) finds that institutional trust enables political participation, especially among less resourceful citizens. Other studies show that trust increases when institutions are seen as inclusive and responsive: Mann et al. (2022) report higher trust among sexual minorities in countries with equal rights legislation; Wilkes and Wu (2019) and Tyrberg (2023) observe similar patterns among ethnic minorities and immigrants.

These findings underscore that trust is shaped not only by institutional performance, but also by perceptions of fairness and representation. To conceptualize this relationship, we introduce the notion of grievance-based trust erosion—a framework explaining how trust deteriorates when individuals accumulate experiences of perceived injustice or exclusion. Grievances may arise from policy outcomes, discriminatory treatment, or broader feelings of marginalization. In contrast to performance-based models, this perspective foregrounds emotional and moral judgments: when fairness norms appear violated, perceived legitimacy declines—even if institutional performance remains unchanged.

Supporting this view, Levy and Sacks (2009) show that perceived government unresponsiveness among disadvantaged groups is associated with lower trust and political efficacy. Anderson and Tverdova (2003) and Oskooii (2024) similarly find that perceived injustice, whether interpersonal or institutional, generates cognitive and emotional responses that undermine confidence in state institutions.

While this framework has typically focused on ethnic or cultural exclusion (Wilkes, 2018; Tyrberg, 2023), political exclusion—based on one's beliefs or partisan identity—has received comparatively less attention. Yet, discrimination or harassment for political reasons may function in similar ways. Such experiences act as signals of exclusion from public discourse or decision-making, leading individuals to feel marginalized from the political community. This sense of exclusion can cause individuals to question the impartiality, responsiveness, and integrity of institutions.

This leads to our first hypothesis:

\emph{\textbf{\small
H1: Political discrimination or harassment has a negative effect on trust in institutions.
}}

Importantly, the effects of political discrimination may not be uniform or isolated from the complexities of other political phenomena. In contexts of rising affective polarization, where political identities become emotionally charged and intergroup hostility intensifies (Iyengar et al., 2019; McCoy et al., 2018), perceptions of fairness and legitimacy are increasingly filtered through partisan lenses.

In this context, we argue that the effects of experiencing political discrimination or harassment on institutional trust are differentiated depending on two key factors: political dissonance and political alignment.

We define political dissonance as the psychological and emotional tension experienced by individuals who perceive a mismatch between their political identity and the dominant political climate or institutional stance (Lelkes, 2016; Federico et al., 2019). Unlike affective polarization, which operates at the societal level, political dissonance is individually experienced and marked by feelings of alienation, frustration, and disillusionment.

Political dissonance is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common (Finkel et al., 2020). This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal (Hetherington \& Rudolph, 2015). In turn, this reduces trust and discourages political participation (Michelson, 2003; Lelkes \& Westwood, 2017).

Accordingly, we propose:

\emph{\textbf{\small
H2: People with different levels of political dissonance respond differently to political discrimination, leading to varying degrees of declining trust in public institutions.
}}

The second mechanism concerns the politicization of institutional perception—the tendency to evaluate institutions based not on their procedures or performance, but on perceived political alignment. In highly polarized contexts, even neutral institutions are judged through partisan filters (Rogowski \& Sutherland, 2015), leading to a fragmentation of legitimacy. Trust becomes conditional on whether individuals perceive institutions as aligned with their own political views.

In such environments, those who experience political discrimination may interpret institutions—even those uninvolved in the discrimination—as biased or complicit. Conversely, individuals aligned with dominant political forces may continue to express trust in the same institutions, despite evidence of exclusion.

This asymmetric trust dynamic reflects the erosion of institutions as shared reference points. Studies show that even traditionally impartial bodies—courts, electoral commissions, public broadcasters—become objects of distrust when seen as favoring one side (Druckman \& Levendusky, 2019; Keefer \& Scartascini, 2022). As Oskooii (2016) notes, perceived exclusion reshapes institutional evaluations, reframing institutions as structurally biased rather than neutral arbiters.

We therefore propose:

\emph{\textbf{\small
H3: Individuals who experience political discrimination will exhibit different patterns of institutional (dis)trust based on their political alignment.
}}

Figure xx presents our conceptual framework, linking political discrimination, affective polarization, and institutional trust through sequential and interactive pathways. At its core is the idea that political discrimination functions as an initial grievance—eroding perceived fairness and inclusion, and in turn diminishing trust in institutions.

In polarized environments, this grievance deepens. Political disagreement becomes moral antagonism, and institutions are increasingly interpreted through partisan lenses. Even neutral bodies are recast as political actors, distorting legitimacy and undermining the institutional consensus on which democratic stability relies.

Ultimately, the interaction between political discrimination and polarization does not produce a uniform decline in trust. Instead, it fragments perceptions of legitimacy. Those who feel politically marginalized are more likely to view institutions as biased or adversarial, while those who feel politically represented may retain confidence. These mechanisms reinforce one another, compounding democratic disaffection, deepening polarization, and threatening the foundations of pluralistic governance.

@@

Due to missing values across the variables of interest, the effective analytical sample for this study is smaller than the total sample published by The World Justice Project. Table \ref{table:1} presents descriptive statistics for the key demographic and political variables among survey respondents included in the study \emph{full} sample.

@@

Researchers typically turn to randomized controlled experiments—the gold standard for causal inference—to address such questions. However, the nature of our research question makes it both practically difficult and ethically problematic to randomly expose individuals to discrimination or harassment based on their political views. As a result, we adopt quasi-experimental designs to examine the causal effects of political discrimination and harassment on institutional trust.

@@

Finally, we investigate the mechanisms through which political discrimination and harassment affect trust in political institutions. We explore two potential pathways: political dissonance and political alignment.

Regarding political polarization, we hypothesize that individuals who experience D/H events may become more politically polarized, which in turn could lead to lower levels of trust in political institutions. To test this hypothesis, we include a measure of political dissonance in our regression model and assess its interaction with the D/H event variable..

%and assess the transmission mechanism using the mediation analysis proposed by Imai et al. (2011) and Imai and Yamamoto (2013).\footnote{We use the mediation R package developed by Tingley et al. (2014) for the practical implementation of this framework.}

%The mediation analysis follows a two-step process. First, we examine how D/H experiences affect political polarization (see Equation \ref{eq:2}). Second, we analyze how both D/H experiences and political polarization jointly influence trust in political institutions (see Equation \ref{eq:3}). We measured political polarization at the individual level using a measure of political conformity/dissonance based on the political ideologies of the respondents. This measure is calculated as the square of the difference between the respondent's political ideology and the average political ideology of the country.\footnote{It is important to note that we use the full study sample to estimate the average political ideology of the country, rather than the matched sample. We believe that using the full representative country sample captures the true distance between the respondent's political ideology and the average political ideology of the country.}

% \begin{equation}
% \label{eq:2}
% M_{ic} = \alpha + \phi X_{ic} + \mathbf{Z}_{ic}'\mathbf{\beta} + \gamma_c + \epsilon_{ic}
% \end{equation}
% \begin{equation}
% \label{eq:3}
% Y_{ic} = \alpha + \delta X_{ic} + \rho M_{ic} + \mathbf{Z}_{ic}'\mathbf{\eta} + \lambda_c + \mu_{ic}
% \end{equation}

% Where:
% \begin{itemize}
%   \item \( Y_{ic} \): political trust for individual \( i \) in region \( c \)
%   \item \( X_{ic} \): D/H event (1 = treated, 0 = control)
%   \item \( M_{ic} \): political dissonance for individual \( i \) in region \( c \)
%   \item $\phi$, $\delta$, and $\rho$: coefficients of interests
%   \item \( \mathbf{Z}_{ic} \): vector of covariates
%   \item $\mathbf{\beta}$ and $\mathbf{\eta}$: coefficient vectors for covariates
%   \item $\gamma_c$ and $\lambda_c$: region fixed effects
%   \item $\epsilon_{ic}$ and $\mu_{ic}$: error terms
% \end{itemize}

% After fitting these models, we use bootstrapping to estimate causal effects under counterfactual scenarios. In each simulation, we compare predicted trust levels between two scenarios: experiencing D/H events versus not experiencing them, while allowing political dissonance to adjust to its expected level under each condition. Averaging across simulations yields the average causal mediation effect (ACME), which captures the indirect effect through polarization, and the average direct effect (ADE), which captures the remaining direct impact.

@@

\begin{equation}
  \label{eq:4}
  log(Y_{ic}) = \alpha + \tau X_{ic} + \delta D_{ic} + \phi X_{ic}D_{ic} + \mathbf{Z}_{ic}'\mathbf{\beta} + \gamma_c + \varepsilon_{ic}
  \end{equation}

  Where:
  \begin{itemize}
    \item \( Y_{ic} \): political trust for individual \( i \) in region \( c \)
    \item \( X_{ic} \): D/H event (1 = treated, 0 = control)
    \item \( D_{ic} \): pathway (high political dissonance or political aligned) for individual \( i \) in region \( c \)
    \item \( \tau \): treatment effect (ATT)
    \item \( \delta \): pathway unilateral effect on Trust in Political Institutions Index
    \item \( \phi \): interaction effect between D/H event and political alignment with the incumbent party
    \item \( \mathbf{Z}_{ic} \): vector of covariates
    \item \( \mathbf{\beta} \): coefficient vector for covariates
    \item \( \gamma_c \): region fixed effects
    \item \( \varepsilon_{ic} \): error term
  \end{itemize}

@@

% \subsection{Polarization as a transmission channel}

% As mentioned earlier, we hypothesize that political polarization serves as a transmission channel through which political discrimination and harassment affect trust in political institutions. To test this hypothesis, we conduct a mediation analysis using the framework developed by Imai et al. (2011) and Imai and Yamamoto (2013). This analysis aims to determine how much of the total effect of D/H experiences on institutional trust operates through political polarization. The framework is based in the estimation of the system presented by equations \ref{eq:2} and \ref{eq:3} along with a bootstrapping procedure to test the mediation channel.

% Table \ref{table:6} presents three key estimates: the total effect, the average causal mediation effect (ACME), and the average direct effect (ADE). As we can observe, experiencing political discrimination or harassment leads to a large and highly significant drop in institutional trust: the Total Effect of –0.863 (on a 0–10 scale) tells us that, on average, those who report discrimination score 0.863 points lower in trust than those who do not. This effect is equivalent to the 18.9\% decrease presented in Table \ref{table:4}, column (III).

% To better understand the mechanisms behind this decline in trust, the analysis separates the effect into two components: the part that operates indirectly through increased political dissonance (the Average Causal Mediation Effect, or ACME), and the part that operates directly, independently of dissonance (the Average Direct Effect, or ADE).

% The estimated ACME is –0.0063, meaning that only about 0.006 points of the decline in trust can be attributed to a pathway where discrimination increases political dissonance, which in turn lowers trust. By contrast, the ADE is –0.857, nearly identical to the total effect, which shows that the vast majority of the impact of discrimination on trust occurs directly, without being mediated by political dissonance.

% In other words, less than 1\% of the total effect of political discrimination on institutional trust can be explained by changes in political dissonance. Although this indirect effect is statistically significant, it is substantively negligible. This implies that political dissonance does act as a mechanism, but it plays only a very minor role. Most of the loss in trust stems from other pathways—perhaps emotional responses like anger or fear, cognitive assessments of systemic injustice, or broader alienation from the political system that are not captured by the dissonance variable.

% \input{"tabs/mediation_analysis.tex"}

\subsection{Heterogenous effects}

@@

The empirical findings provide robust support for the central premise of this study: experiencing political discrimination significantly erodes institutional trust, affirming our first hypothesis (H1). This result suggests that political exclusion is not a marginal experience but a core driver of institutional delegitimation. Individuals who report discrimination or harassment based on their political beliefs express substantially lower levels of trust in public institutions.

These findings validate our theory of grievance-based trust erosion, demonstrating that institutional trust is shaped not only by objective outcomes or policy performance but also by perceived violations of fairness and inclusion. When individuals feel politically targeted, these experiences evoke emotional and moral reactions that lead them to reinterpret institutions as biased or adversarial. In this sense, political discrimination functions as a distinct form of grievance—one that undermines institutional legitimacy from the bottom up.

By centering political exclusion, this study extends prior research focused on ethnic and minority-based grievances (Wilkes, 2018; Tyrberg, 2023) to the domain of political and partisan identity, showing that subjective perceptions of injustice based on political beliefs are equally corrosive to democratic legitimacy. This underscores the need to consider how institutional trust is not only eroded by what governments do, but also by how citizens experience their political belonging within the system.

In politically polarized societies—where party identities are emotionally charged and social divisions are entrenched—the effect of political discrimination on trust unfolds through two distinct but interrelated cognitive mechanisms.

The first is political dissonance, which occurs when individuals feel ideologically misaligned with the dominant political climate. Our findings support the second hypothesis (H2): individuals who both experience political discrimination and feel ideologically out of place exhibit the most pronounced declines in trust. This sense of disconnection intensifies perceptions of exclusion, reinforcing the belief that the political system no longer represents them (McCoy et al., 2018; Finkel et al., 2020).

As dissonance increases, experiences of discrimination are no longer perceived as isolated events but as symptomatic of a broader political environment seen as hostile. This magnifies resentment, sharpens perceived injustice, and increases receptiveness to populist or anti-establishment narratives that portray institutions as fundamentally biased. In this way, affective polarization amplifies the emotional and psychological toll of exclusion, deepening distrust and weakening democratic attachment.

The second mechanism is politicized perception (Rogowski \& Sutherland, 2015; Druckman \& Levendusky, 2019). In highly polarized contexts, citizens evaluate institutions not based on procedural fairness or performance, but through partisan filters. Our findings support hypothesis three (H3): political discrimination contributes to the reframing of institutions—even neutral ones—as partisan actors.

This mechanism differs from dissonance in that it focuses not on ideological mismatch, but on a shift in how institutions are perceived. Rather than being viewed as impartial guarantors of democracy, institutions are interpreted as aligned with opposing political forces. For individuals who have experienced political discrimination, this cognitive shift is especially pronounced. Prior exclusion fosters suspicion, increasing the likelihood that institutional behavior is seen as adversarial or punitive.

Affective polarization further reinforces this interpretive lens. As institutions are increasingly judged by their perceived political alignment, trust erodes—not just because of what institutions do, but because of what they are perceived to represent. This dynamic undermines not only confidence in specific institutions, but the very notion of institutional neutrality, weakening the symbolic and procedural bonds that tie citizens to the democratic state.

Taken together, these findings reveal a layered and interactive erosion of institutional trust in polarized political environments. Rather than operating through a single channel, the effects of political discrimination unfold through mutually reinforcing cognitive pathways. Political dissonance highlights the emotional and psychological costs of ideological alienation, while politicized perception reshapes how individuals interpret the role and legitimacy of institutions.

These processes fragment the shared symbolic space in which institutions are recognized as legitimate by all citizens, regardless of ideology. In doing so, they weaken the connective tissue that binds democratic institutions to the public they serve.

Crucially, this erosion is not a standalone process. It is embedded within broader structural and contextual conditions—marked by sustained polarization, group antagonism, and diminishing consensus over institutional roles. In such environments, institutions are increasingly viewed through affective and partisan lenses, shifting evaluations of legitimacy from questions of fairness and procedure to perceptions of political alignment. As a result, trust becomes contingent, and the normative ideal of nonpartisan institutional authority begins to unravel.

Understanding these interlocking processes is essential not only for assessing democratic resilience, but also for identifying the pressure points at which institutional legitimacy becomes most vulnerable to collapse.
